---
title: "Assignment 2: Coding Basics"
author: "Erchi Luo"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons/labs in Environmental Data Analytics on coding basics.

## Directions

1.  Rename this file `<FirstLast>_A02_CodingBasics.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction.
4.  Be sure to **answer the questions** in this assignment document. 
5.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
6.  After Knitting, submit the completed exercise (PDF file) to Canvas.

7.  Initial here to acknowledge that you did not use AI at all in completing this assignment: Erchi Luo


## Basics, Part 1

1.  Use R's `seq()` function to create a sequence of numbers from 100 to 333, increasing by threes. Assign this sequence a variable name.
2.  Compute the *mean* of this sequence, assigning this values its own variable name. 
3.  Compute the *standard deviation* (`sd()`) of this sequence, assigning this values its own variable name.  
4.  Display the the mean minus the standard deviation as well as the mean plus the standard deviation.
5.  Insert comments in your code to describe what you are doing.

```{r basics 1}
#1. Create a sequence from 100 to 333, increasing by 3s
my_sequence <- seq(from = 100, to = 333, by = 3)

#2. Compute the mean of this sequence and assign it to a variable name
sequence_mean <- mean(my_sequence)

#3. Compute the standard deviation of this sequence and assign it to a variable name
sequence_sd <- sd(my_sequence)

#4. Display the mean minus the standard deviation and the mean plus the standard deviation
sequence_mean - sequence_sd
sequence_mean + sequence_sd

```

---

## Basics, Part 2
6.  Create three vectors, each with four components, consisting of (a) student names, (b) test scores, and (c) whether they are on scholarship or not (TRUE or FALSE).

7.  Label each vector with a comment on what type of vector it is.

8.  Combine each of the vectors into a data frame. Assign the data frame an informative name.

9.  Label the columns of your data frame with informative titles.

```{r basics 2}
# 6 & 7. Create three vectors with four components and label their types
# Character vector for student names
student_names <- c("James", "Curry", "Durant", "Hansen")
# Numeric vector for test scores
test_scores <- c(88, 92, 95, 81)
# Logical vector for scholarship status (TRUE or FALSE)
on_scholarship <- c(TRUE, FALSE, FALSE, TRUE)
# 8 & 9. Combine vectors into a data frame with informative column titles
student_records <- data.frame(
  "Student_Name" = student_names, 
  "Final_Score" = test_scores, 
  "Has_Scholarship" = on_scholarship
)
# Display the data frame to check the result
student_records

```

10.  QUESTION: How is this data frame different from a matrix?

> Answer:
A data frame can contain columns with different data types (e.g., characters, numbers, and logical values), while a matrix requires all elements to be of the same data type.
---

## Basics, Part 3

11. Create a function with one input. In this function, use `if`...`else` to evaluate the value of the input: if it is greater than 50, it returns the word "Pass"; otherwise it returns the word "Fail". 

12. Create a second function that does the exact same thing as the previous one but uses `ifelse()` instead if `if`...`else `. 

13. Run both functions using the value 54 as the input

14. Run both functions using the **vector** of student test scores you created as the input. (Only one will work properly...)

```{r basics 3}
#11. Create a function using if...else
# 11. Create a function using if...else
# This function handles a single input value
check_score_if <- function(score) {
  if (score > 50) {
    return("Pass")
  } else {
    return("Fail")
  }
}
#12. Create a function using ifelse()
check_score_ifelse <- function(score) {
  result <- ifelse(score > 50, "Pass", "Fail")
  return(result)
}
#13a. Run the first function with the value 54
check_score_if(54)
#13b. Run the second function with the value 54
check_score_ifelse(54)
#14a. Run the first function with the vector of test scores
# Note: This will likely generate a warning because if() only checks the first element
#check_score_if(test_scores)
#14b. Run the second function with the vector of test scores
# This will work correctly for every score in the vector
check_score_ifelse(test_scores)
```

15. QUESTION: Which option of `if`...`else` vs. `ifelse` worked? Why? (Hint: search the web for "R vectorization"; it's ok here if an AI response is presented in the search response.)

> Answer:The ifelse option worked correctly for the vector, while the if...else function generated an error. This is because if statements in R are not vectorized and can only evaluate one logical value at a time. In contrast, ifelse() is a vectorized function designed to evaluate every element in a vector simultaneously. This concept is known as "R Vectorization."(I searched this answer through AI.)


**NOTE** Before knitting, you'll need to comment out the call to the function in Q14 that does not work. (A document can't knit if the code it contains causes an error!)