---
title: "Assignment 3: Data Exploration"
author: "Erchi Luo"
date: "Spring 2026"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  [NEW] Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to Canvas.

8.  Initial here to acknowledge that you did not use AI in completing this assignment, except where expressly allowed: _Erchi Luo__

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks in your code chunks.

**TIP**: If your code fails to knit, check:
 * That no `install.packages()` or `View()` commands exist in your code. 
 * That you are not displaying the entire contents of a large dataframe in your code.

---

## Set up your R session

1.  Load necessary packages (tidyverse, here), check your current working directory and import two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. 

**Be sure to**:
* Use the `here()` package in specifying the paths to your datasets
* Include the appropriate subcommand to read in character based columns as factors


```{r,  message = FALSE}
# 1. Load the necessary packages
library(tidyverse)
library(here)
# 2. Check the working directory using the here package
getwd()
here()
#3. Import datasets
# I define the path to include both "Data" and "Raw" folders
data_path <- "Data/Raw"

# Import Neonics dataset from the Raw folder
Neonics <- read.csv(
  here(data_path, "ECOTOX_Neonicotinoids_Insects_raw.csv"), 
  stringsAsFactors = TRUE
)

# Import Litter dataset from the Raw folder
Litter <- read.csv(
  here(data_path, "NEON_NIWO_Litter_massdata_2018-08_raw.csv"), 
  stringsAsFactors = TRUE
)
```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer: Well, I think we are interested in this because neonicotinoids don't just kill the 'bad' pests; they also hit beneficial insects like bees and butterflies. Since these pollinators are crucial for our food supply, we need to understand how much of this chemical they can handle before it starts wiping out their populations.


3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information. (AI is allowed here, but put answers in your own words.)

> Answer:From the "Climate Change & Carbon" perspective.Litter and woody debris act like a 'blanket' for the forest floor. They help hold moisture in the soil, prevent erosion, and provide a home for insects and fungi. Also, because this debris stores a lot of carbon, studying it helps us understand how forests are storing greenhouse gases instead of releasing them into the atmosphere.


4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Trap Types and Placement: NEON uses two distinct types of traps to capture different debris sizes. Elevated traps (0.5 m² square mesh "baskets" elevated about 80 cm off the ground) are used to collect litter and smaller debris, while ground traps (3 m x 0.5 m rectangular areas) are used to sample larger pieces of fine woody debris that aren't reliably caught by elevated traps.
 2.Variable Sampling Frequency: The timing for collection depends on the type of forest. In deciduous forests, sampling is concentrated during fall senescence (when leaves drop), while in evergreen and coniferous forests, sampling occurs year-round, typically at monthly intervals.
 3. Functional Group Sorting: Once collected, all material is sorted into specific functional groups—such as leaves, needles, twigs/branches, seeds, and flowers—before being dried and weighed to determine the mass of each component.


## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied.  [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
effect_counts <- summary(Neonics$Effect)
sort(effect_counts, decreasing = TRUE)
```

> Question: Which two effects stand out as the most studied? Can you guess why these effects might specifically be of interest?
> Answer:The two effects that stand out as the most studied are Population (1803 observations) and Mortality (1493 observations).These are likely of specific interest because they are the most direct indicators of a chemical's impact on a species. Mortality is a critical metric for determining the immediate lethality of an insecticide, while Population effects show how the chemical influences the long-term survival and growth of the entire insect community, which is essential for understanding broader ecological consequences.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name).[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
species_summary <- summary(Neonics$Species.Common.Name, maxsum = 10)
sort(species_summary, decreasing = TRUE)

```

> Question: What do these species have in common? Why might they be of interest over other insects? 
> Answer:What these species have in common is that the vast majority of them are pollinators, specifically different types of bees (Honey Bees and Bumble Bees). Even the Parasitic Wasp plays a significant role in ecosystems as a natural form of pest control.
These species are likely of greater interest than other insects because of their economic and ecological importance. Bees are essential for pollinating many of the crops that make up our food supply.


8.  The `Conc.1..Author` column, which lists the concentration of the neonicitoid dose, should include numeric values. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
# Check the class of the concentration column
class(Neonics$Conc.1..Author)

```

> Answer:The class of the Conc.1..Author column is "Factor". It is not numeric because the original dataset likely contains non-numeric characters within that column.


## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
Neonics$Publication.Year <- as.numeric(as.character(Neonics$Publication.Year))
ggplot(Neonics, aes(x = Publication.Year)) +
  geom_freqpoly(binwidth = 1) +
  labs(
    title = "Number of Neonicotinoid Studies Over Time",
    x = "Publication Year",
    y = "Number of Studies"
  )

```


10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics, aes(x = Publication.Year, color = Test.Location)) +
  geom_freqpoly(binwidth = 1) +
  labs(
    title = "Neonicotinoid Studies Over Time by Test Location",
    x = "Publication Year",
    y = "Number of Studies",
    color = "Test Location"
  ) +
  theme_minimal()
```


>Interpret this graph. What are the most common test locations, and do they differ over time?
> Answer: According to the graph, "Lab" is the most common test locations after 2010.In some years before 2010, "Field natural" would have been more popular.


11. Create a bar graph of Endpoint counts. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
ggplot(Neonics, aes(x = Endpoint)) +
  geom_bar() +
  labs(
    title = "Frequency of Different Study Endpoints",
    x = "Endpoint",
    y = "Number of Studies"
  ) +
  # use the tip
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```


> What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix (p.721) for more information. 
> Answer: LOEL and NOEL.

---

## Explore your data (Litter)

12. Determine the class of `collectDate`. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
Litter$collectDate <- as.Date(Litter$collectDate)
class(Litter$collectDate)

august_dates <- unique(Litter$collectDate[format(Litter$collectDate, "%m-%Y") == "08-2018"])
print(august_dates)
```


13. Using the `unique` function, list the different `plotIDs` sampled at Niwot Ridge. 

```{r}
niwo_plots <- unique(Litter$plotID[Litter$siteID == "NIWO"])
print(niwo_plots)
```

> How is the information obtained from `unique` different from that obtained from `summary`?
> Answer:The difference lies in identity versus distribution: unique() provides a list of distinct values in a column. It tells you what the different categories or entries are (e.g., a list of specific Site names), removing all duplicates.summary() provides a statistical overview of the column.For numeric data, it gives the min, max, mean, and quartiles.For categorical data (Factors), it gives the count (frequency) of each unique value.

14. Create a bar graph of `functionalGroup` counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}
library(ggplot2)
library(dplyr)

# Step 1: Filter the data to select entries where the siteID is "NIWO".
Litter_NIWO <- filter(Litter, siteID == "NIWO")

# Step 2: Create a bar chart.
ggplot(Litter_NIWO, aes(x = functionalGroup)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Litter Functional Groups at Niwot Ridge",
    x = "Functional Group",
    y = "Number of Samples"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of `dryMass` by `functionalGroup`.

```{r}
ggplot(Litter, aes(x = functionalGroup, y = dryMass)) +
  # 1. The Violin Plot shows the density/shape of the data
  geom_violin(fill = "lightblue", alpha = 0.4, na.rm = TRUE) +
  # 2. The Boxplot shows the median and quartiles
  # We make it skinny (width = 0.1) so it fits inside the violin
  geom_boxplot(width = 0.1, color = "black", outlier.size = 0.5, na.rm = TRUE) +
  # 3. Use coord_cartesian to "zoom in" if outliers are too big
  # (Optional: remove the '#' below if your graph looks too flat)
  # coord_cartesian(ylim = c(0, 20)) + 
  labs(
    title = "Dry Mass Distribution by Functional Group",
    x = "Functional Group",
    y = "Dry Mass (grams)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

> Why is the boxplot a more effective visualization option than the violin plot in this case?
> Answer:In this case, the boxplot is more effective because the dryMass data is highly skewed with significant outliers. The boxplot clearly identifies these outliers as individual points and provides a precise visual representation of the median and quartiles, which are more useful for comparing the biomass of different functional groups than the density shape provided by a violin plot.


> What type(s) of litter tend to have the highest biomass at these sites?
> Answer:Needles.
