---
title: "Assignment 8: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Erchi Luo"
date: "Spring 2026"
output: pdf_document
latex_engine: xelatex
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A08_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.

**!! AI generated code should NOT be used in this assignment**

>NOTE: For this assignment, you can skip tests for any of the assumptions required to run the models. 

## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1 Set up session
# Load needed packages
library(tidyverse)
library(lubridate)
library(here)
library(agricolae)

# Import raw data
ntl_data <- read.csv(
  here("Data", "Raw", "NTL-LTER_Lake_ChemistryPhysics_Raw.csv"), 
  stringsAsFactors = TRUE
)

# Set date columns to date objects
ntl_data$sampledate <- mdy(ntl_data$sampledate)

# Confirm the date conversion was successful
class(ntl_data$sampledate)

#2 Custom ggplot theme
# Build my own theme
my_theme <- theme_classic() + 
  theme(
    line = element_line(color = "black", linewidth = 0.5),
    legend.position = "right",
    legend.title = element_text(size = 12),
    plot.title = element_text(color = "darkblue", size = 14, face = "bold"),
    axis.text = element_text(size = 10, color = "black"),
    axis.title = element_text(size = 12),
    panel.grid.major = element_line(color = "grey90"),
    panel.grid.minor = element_blank()
  )
# Set it as the default theme for the entire session
theme_set(my_theme)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There is no significant relationship between lake depth and mean lake temperature in July. 
Ha: There is a significant relationship between lake depth and mean lake temperature in July. 

4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4 Wrangle data for July temperature analysis
ntl_july_processed <- ntl_data %>%
  # Filter for records collected only in July
  filter(month(sampledate) == 7) %>%
  
  # Select only the required columns
  select(lakename, year4, daynum, depth, temperature_C) %>%
  
  # Keep only complete cases (remove rows with NAs)
  drop_na()

# Check the dimensions and the first few rows of the processed data
dim(ntl_july_processed)
head(ntl_july_processed)

#5 Visualizing Temperature vs. Depth
# Create the scatter plot
temp_depth_plot <- ggplot(ntl_july_processed, aes(x = depth, y = temperature_C)) +
  # Add points with some transparency (alpha) to handle overplotting
  geom_point(alpha = 0.3, color = "darkblue") +
  # Add the linear regression line
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  # Set y-axis limits from 0 to 35 °C
  ylim(0, 35) +
  # Label the plot
  labs(
    title = "Relationship between Lake Depth and Temperature (July)",
    subtitle = "Data source: NTL-LTER",
    x = "Depth (m)",
    y = "Temperature (°C)"
  )

# Display the plot
print(temp_depth_plot)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The plot suggests a strong negative correlation between lake depth and temperature during July. As depth increases, the temperature consistently decreases.
Linearity of the trend: While the red regression line shows a downward linear trend, the distribution of the points suggests that the relationship may not be perfectly linear across the entire range.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7 Perform linear regression
# Build the linear model: temperature as a function of depth
temp_depth_lm <- lm(temperature_C ~ depth, data = ntl_july_processed)

# Display the summary results
summary(temp_depth_lm)


```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer:The linear regression model indicates a statistically significant relationship between lake depth and water temperature in July (p < 0.001).Variance Explained: The model explains approximately 73.9% of the variability in temperature (R^2 = 0.7387).Degrees of Freedom: This finding is based on 9,726 degrees of freedom (residual degrees of freedom), indicating a robust analysis supported by a large sample size.Predictive Change: For every 1-meter increase in depth, the model predicts a 1.95 °C decrease in water temperature (slope = -1.946).


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9 AIC for model selection
# Load the 'MASS' package for the stepAIC function
library(MASS)

# Create a "full" model with all potential predictors
full_model <- lm(temperature_C ~ depth + year4 + daynum, data = ntl_july_processed)

# Run stepAIC to find the best model
# direction = "backward" starts with full model and removes least useful variables
best_model_aic <- stepAIC(full_model, direction = "backward", trace = FALSE)

# Display the results to see which variables are selected
summary(best_model_aic)


#10 Run the multiple regression model
# Based on AIC, we use all three predictors
final_multi_model <- lm(temperature_C ~ depth + year4 + daynum, data = ntl_july_processed)

# Display the summary to interpret the effect of each predictor
summary(final_multi_model)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The AIC method suggests using all three variables (depth, year4, and daynum) to predict lake temperature, as they all contribute significantly to the model.This multiple regression model explains approximately 74.1% of the observed variance in lake temperature (R^2 = 0.7412).
Comparison with the simple model: This is an improvement over the model using only depth as an explanatory variable , the inclusion of year4 and daynum provides a more robust model that accounts for temporal trends and seasonal variations within the monitoring period.


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12 ANOVA and Linear Model for Lake Temperatures
# ANOVA model using aov()
anova_model <- aov(temperature_C ~ lakename, data = ntl_july_processed)

# Linear model using lm()
lm_model <- lm(temperature_C ~ lakename, data = ntl_july_processed)

# Display the results
summary(anova_model)

# View detailed parameters of the linear model
summary(lm_model)

```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Yes, there is a highly significant difference in mean water temperatures among the different lakes (p < 2.2e-16).Findings: The linear model confirms that lakename is a significant predictor of water temperature. The model coefficients indicate that most lakes differ significantly from the reference lake (the intercept group).



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14. Scatter plot of Temperature by Depth, grouped by Lake
# Create the visualization
lake_temp_plot <- ggplot(ntl_july_processed, aes(x = depth, y = temperature_C, color = lakename)) +
  # Add points with 50% transparency
  geom_point(alpha = 0.5) +
  # Add a linear regression line for each lake
  geom_smooth(method = "lm", se = FALSE) +
  # Set y-axis limits
  ylim(0, 35) +
  # Labeling and styling
  labs(
    title = "Temperature by Depth across Different Lakes (July)",
    x = "Depth (m)",
    y = "Temperature (°C)",
    color = "Lake Name"
  ) +
  theme_bw() + # Clean, professional theme
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold")
  )

# Display the plot
print(lake_temp_plot)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15 Perform Tukey's HSD test
# Run the Tukey HSD test on the ANOVA model
tukey_results <- TukeyHSD(anova_model)

# To keep the output clean, we can look at the first few comparisons
head(tukey_results$lakename)

# Visualization is often better than a long list of text
# This plots the differences in means and their confidence intervals
par(mar = c(5, 15, 2, 2), mgp = c(3, 1, 0))
plot(tukey_results, las = 1, cex.axis = 0.5, col = "blue")

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:Since the confidence interval for this pair crosses the 0 line, we can conclude that Paul Lake has a mean temperature that is statistically indistinguishable from Peter Lake.East Long Lake appears to be the most distinct. In the plot, many of its comparison intervals are shifted far to the left or right of the 0 line and do not cross it when compared to several other lakes.Conversely, Central Long Lake also exhibits wide separation in many of its comparisons.

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: two-sample t-test.



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
#18 T-test for Crampton Lake and Ward Lake
# Filter the data for the two specific lakes
crampton_ward_july <- ntl_july_processed %>%
  filter(lakename == "Crampton Lake" | lakename == "Ward Lake")

# Run a two-sample t-test
# We assume unequal variance (Welch's T-test) which is the default in R
lake_t_test <- t.test(temperature_C ~ lakename, data = crampton_ward_july)

# Display results
print(lake_t_test)
```

>Answer: The Welch Two Sample t-test results (t = 1.1181, df = 200.37, p = 0.2649) indicate that there is no statistically significant difference in the mean July temperatures between Crampton Lake and Ward Lake.
The mean temperatures are considered equal.Since the p-value is greater than the standard alpha level of $0.05$ and the 95% confidence interval includes zero ([-0.682, 2.469]), we fail to reject the null hypothesis.This matches the answer for Part 16.

