---
title: "Assignment 10: Data Scraping"
author: "Erchi Luo"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 Set up session
# Load necessary packages
# tidyverse for data manipulation, rvest for web scraping
library(tidyverse)
library(rvest)
library(lubridate) 
library(here)      

# Check and display the current working directory
getwd()

# Alternatively, verify the project root using the here package
here()

```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2024 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Set the year to 2024
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 
# Assign the target URL to a variable
url <- "https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2024"

# Read the HTML content of the webpage into an rvest object
# This stores the entire structure of the page (tags, text, tables) in 'durham_2024_webpage'
durham_2024_webpage <- read_html(url)

# Print the object to confirm it loaded correctly
print(durham_2024_webpage)

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * Contact Person
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "Reginald Hicks", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 Scrape Specific System and Water Use Data
# Scrape Water System Name (Expected: "Durham")
# We use str_trim to remove any leading/trailing whitespace or newlines
water_system_name <- durham_2024_webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text() %>% 
  str_trim()

# Scrape Contact Person (Expected: "Reginald Hicks")
contact_person <- durham_2024_webpage %>% 
  html_nodes("div+ table tr:nth-child(4) td:nth-child(2)") %>% 
  html_text() %>% 
  str_trim()

# Scrape Ownership (Expected: "Municipality")
ownership <- durham_2024_webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text() %>% 
  str_trim()

# Scrape Maximum Day Use (MGD) for each month
# This should result in a vector of 12 character strings
max_day_use <- durham_2024_webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text() %>% 
  str_trim()

# --- Verification Step ---
# Print the variables to the console to ensure they match the requirements
print(paste("System Name:", water_system_name))
print(paste("Contact:", contact_person))
print(paste("Ownership:", ownership))

# Check the vector of 12 monthly values
print("Monthly Maximum Day Use (MGD):")
print(max_day_use)

# Verify that we have exactly 12 values for the months
length(max_day_use)


```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... 

5. Create a line plot of the maximum daily withdrawals across the months for 2024, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 Convert Scraped Data into a Dataframe
# a. Define the Year
scraped_year <- "2024"

# b. Define the months in the order they appear on the website (row-wise)
# Based on the website layout: Jan, May, Sep are in the first row of data
month_order <- c("Jan", "May", "Sep", 
                 "Feb", "Jun", "Oct", 
                 "Mar", "Jul", "Nov", 
                 "Apr", "Aug", "Dec")

# c. Assemble the dataframe
# We use rep() to repeat the single values 12 times each
durham_lwsp_2024 <- data.frame(
  "System_Name" = rep(water_system_name, 12),
  "Contact_Person" = rep(contact_person, 12),
  "Ownership" = rep(ownership, 12),
  "Month" = month_order,
  "Year" = rep(scraped_year, 12),
  "Max_Day_Use_MGD" = as.numeric(max_day_use) # Convert strings to numbers
)

# d. Create a proper Date column
# We use lubridate's my() function (month-year) to create a real R date object
durham_lwsp_2024 <- durham_lwsp_2024 %>% 
  mutate(Date = my(paste(Month, Year))) %>% 
  # Arrange by date so the data is finally in chronological order
  arrange(Date)

# --- Final Check ---
# View the first few rows of the new tidy dataframe
head(durham_lwsp_2024)


#5 Plot Maximum Daily Withdrawals for 2024 ---

# Create the line plot
# We map Date to the x-axis to ensure correct chronological sequence
withdrawal_2024_plot <- ggplot(durham_lwsp_2024, aes(x = Date, y = Max_Day_Use_MGD)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "darkblue", size = 2) +
  # Use scale_x_date to ensure the month names appear on the x-axis
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  labs(
    title = "2024 Maximum Daily Water Withdrawals in Durham",
    subtitle = "Source: NC DEQ Local Water Supply Planning",
    x = "Month",
    y = "Maximum Day Use (MGD)"
  ) +
  theme_minimal()

# Display the plot
withdrawal_2024_plot

```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function with two inputs, "PWSID" and "year", that:
  - Creates a URL pointing to the LWSP for that PWSID for the given year
  - Creates a website object and scrapes the data from that object (just as you did above)
  - Constructs a dataframe from the scraped data, mostly as you did above, but includes the PWSID and year provided as function inputs in the dataframe. 
  - Returns the dataframe as the function's output


```{r construct.a.scraping.function}
#6. Create a Scrape Function
scrape_lwsp_data <- function(pwsid, year) {
  
  # a. Construct the URL based on provided inputs
  # We use paste0 to stitch the base URL and the arguments together
  url <- paste0("https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=", 
                pwsid, "&year=", year)
  
  # b. Read the webpage content
  webpage <- read_html(url)
  
  # c. Scrape specific values using the selectors from earlier
  # Using html_node to ensure we only get one value for metadata
  system_name <- webpage %>% 
    html_node("td tr:nth-child(1) td") %>% 
    html_text() %>% 
    str_trim()
  
  contact_person <- webpage %>% 
    html_node("tr:nth-child(4) td") %>% 
    html_text() %>% 
    str_trim()
  
  ownership <- webpage %>% 
    html_node("td~ td+ td") %>% 
    html_text() %>% 
    str_trim()
  
  # Scrape the 12 monthly max daily withdrawal values
  max_day_use <- webpage %>% 
    html_nodes("th~ td+ td") %>% 
    html_text() %>% 
    str_trim()
  
  # d. Define month order (matches website table layout)
  month_order <- c("Jan", "May", "Sep", 
                   "Feb", "Jun", "Oct", 
                   "Mar", "Jul", "Nov", 
                   "Apr", "Aug", "Dec")
  
  # e. Construct the dataframe
  df_lwsp <- data.frame(
    "PWSID" = rep(pwsid, 12),
    "Year" = rep(year, 12),
    "System_Name" = rep(system_name, 12),
    "Contact_Person" = rep(contact_person, 12),
    "Ownership" = rep(ownership, 12),
    "Month" = month_order,
    "Max_Day_Use_MGD" = as.numeric(max_day_use)
  )
  
  # f. Add a proper Date column and sort
  df_lwsp <- df_lwsp %>% 
    mutate(Date = my(paste(Month, Year))) %>% 
    arrange(Date)
  
  # Return the final dataframe
  return(df_lwsp)
}
```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2020. Then show the structure of the dataframe with either the `str()` or the `glimpse()` function. 

```{r fetch.and.plot.Durham.2020.data}
#7 Scrape and Plot 2020 Data for Durham ---
# a. Use the function to extract data for Durham in 2020
# PWSID is '03-32-010', Year is 2020
durham_lwsp_2020 <- scrape_lwsp_data("03-32-010", 2020)

# b. Show the structure of the dataframe
# glimpse() is the tidyverse version of str() and provides a clean summary
glimpse(durham_lwsp_2020)

# c. Plot the max daily withdrawals for 2020
# We reuse the logic from Question 5
withdrawal_2020_plot <- ggplot(durham_lwsp_2020, 
                               aes(x = Date, y = Max_Day_Use_MGD)) +
  geom_line(color = "darkgreen") + # Changed color to distinguish from 2024
  geom_point() +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b") +
  labs(
    title = "2020 Durham Maximum Daily Water Withdrawals",
    subtitle = "Data scraped via scrape_lwsp_data function",
    x = "Month",
    y = "Max Day Use (MGD)"
  ) +
  theme_minimal()

# Render the plot
withdrawal_2020_plot
```

8. Use the function above to extract data for Cary (PWSID = '03-92-020') in 2020. Combine this data with the Durham data collected above and create a plot that compares Cary's to Durham's water withdrawals.

```{r fetch.and.plot.Cary.2020.data}
#8 Compare Cary and Durham 2020 Data
# a. Scrape data for Cary using its PWSID for 2020
cary_lwsp_2020 <- scrape_lwsp_data("03-92-020", 2020)

# b. Combine the Durham 2020 and Cary 2020 dataframes
# bind_rows() stacks the two dataframes on top of each other
combined_lwsp_2020 <- bind_rows(durham_lwsp_2020, cary_lwsp_2020)

# c. Create a comparison plot
# We map 'System_Name' to color to create two distinct lines
comparison_plot <- ggplot(combined_lwsp_2020, 
                          aes(x = Date, 
                              y = Max_Day_Use_MGD, 
                              color = System_Name)) +
  geom_line(linewidth = 1) +
  geom_point() +
  scale_x_date(date_breaks = "1 month", 
               date_labels = "%b") +
  labs(
    title = "Comparison of Max Daily Water Withdrawals (2020)",
    subtitle = "Durham vs. Cary Municipality",
    x = "Month",
    y = "Max Day Use (MGD)",
    color = "Water System"
  ) +
  theme_minimal()

# Display the comparison plot
comparison_plot
```


9. Use the code & function you created above to plot Cary's max daily withdrawal by months for the years 2018 thru 2023. Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "06_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs (where one of the two inputs does not change). Pipe the output of the map2() function to `bind_rows()` to combine the dataframes into a single one, and use that to construct your plot.

```{r fetch.and.plot.Cary.multiyear.data }
#9 Scrape Cary Data (2018-2023) and Plot Trend 
# a. Define the inputs for the map2 function
# We want Cary's PWSID repeated for each year we are scraping
pwsid_cary <- "03-92-020"
years_to_scrape <- seq(2018, 2023) # Creates a vector: 2018, 2019, ..., 2023

# b. Use map2 to run the function over the vector of years
# map2(input1, input2, function)
# we pipe the list output to bind_rows() to get one big dataframe
cary_multi_year_df <- map2(pwsid_cary, years_to_scrape, scrape_lwsp_data) %>% 
  bind_rows()

# c. Create the plot with a smoothed line
cary_trend_plot <- ggplot(cary_multi_year_df, 
                          aes(x = Date, y = Max_Day_Use_MGD)) +
  geom_line(color = "grey70", alpha = 0.6) + # Background lines for raw data
  geom_point(size = 1, alpha = 0.5) +
  # Add the smoothed trend line (LOESS)
  geom_smooth(method = "loess", color = "blue", se = TRUE) +
  labs(
    title = "Cary Max Daily Water Withdrawals (2018-2023)",
    subtitle = "Smoothed trend line represents overall usage patterns",
    x = "Year",
    y = "Max Day Use (MGD)"
  ) +
  theme_minimal()

# Display the final plot
cary_trend_plot
```

Question: Just by looking at the plot (i.e. not running statistics), does Cary have a trend in water usage over time?
> Answer:From the plot, we can see that Cary's water usage has a slight upward trend over the 5-year period. While the usage looks fairly flat between 2018 and early 2021, it starts to climb a bit more noticeably from mid-2021 onwards. 
>
